---
title: 'Day 3: GLMMs'
author: "Claudio D. Silva Jr"
output: html_document
---

# Day 3: Generalized Linear Mixed Models - GLMMs

Learning objectives:

-   What are GLMMs

-   What is a distributional assumption

-   Working with GLMMs

$$
\\[0.2in]
$$

## What are GLMMs

-   Generalized Linear Models are models in which we can assume different distributions for our data beyond the Normal distribution.

-   Similar to general linear models, GLMs can also have random effects, thus, Generalized Linear Mixed Models - GLMMs.

$$
\\[0.2in]
$$

### The structure of a GLMMs

Remember that for a **LMMs**, assuming $\mathbf{y}$ arises from a normal distribution, we have:

$$
\mathbf{y = X\beta + Zu + \epsilon} \\ \mathbf{\begin{bmatrix} u \\ \epsilon \end{bmatrix} \sim \begin{pmatrix}  \begin{bmatrix} 0 \\ 0 \end{bmatrix}, \begin{bmatrix} G \; 0 \\ 0 \; R \end{bmatrix} \end{pmatrix}}
$$

In which:

-   $\mathbf{X\beta}$ represents our fixed effects, where $\mathbf{X}$ is a matrix containing our explanatory variables and $\mathbf{\beta}$ a vector containing the fixed-effects parameters.

-   $\mathbf{Zu}$ represents our random effects, where $\mathbf{Z}$ is a design matrix and $\mathbf{u}$ is the vector containing the random effects parameters.

-   $\mathbf{\epsilon}$ is the vector containing the residuals.

-   From $\mathbf{Zu + \epsilon}$ we have: $\mathbf{G}$ is the variance-covariance matrix of the random effects, and $\mathbf{R}$ is the variance-covariance matrix of the residuals.

    -   $\mathbf{G = \sigma^2_uI}$ and $\mathbf{R = \sigma^2I}$, in which $\mathbf{I}$ is the identity matrix.

Which is similar to:

$$
\mathbf{u} \sim N(0, \mathbf{I}\sigma^2_u) \\
\mathbf{\epsilon} \sim N(0, \mathbf{I}\sigma^2)
$$

In this case:

$$
E(\mathbf{y}) = \mathbf{X\beta}, \\
Var(\mathbf{y}) = \mathbf{ZGZ' + R}
$$

We can also write this model as:

$$
\mathbf{y} \sim N(\mathbf{X\boldsymbol{\beta}}, \; \mathbf{ZGZ' + R})
$$

or:

$$
\mathbf{y} \sim N(\mathbf{X\beta}, \; \mathbf{\Sigma}) \\
\mathbf{\Sigma} = \mathbf{ZGZ' + R}
$$

For **GLMMs** the structure changes based on the distribution we will assume for $\mathbf{y}$, but is very similar to the last notation presented. A generic definition would be:

$$
\mathbf{y|u} \sim P(\mu, \; \phi)
$$

In which:

-   Linear predictor: $g(\mu) = \eta = \mathbf{X\beta + Zu}$

    -   $g(\mu) = \eta$ is the link function applied to the expected value.

    -   $E(\mathbf{y|u}) = \mu$.

$$
\\[0.2in]
$$

### Components of GLMMs

#### Link Functions

Our linear predictor $\mathbf{X\beta}$ can produce all possible values in the y-axis of a plot, from $- \; \infty$ to $+ \; \infty$ depending on the value of the predictor variable. A link function links the linear predictor and the distribution assumed for the data $\mathbf{y}$.

In the **link scale**, the mean of $\mathbf{y}$ respect linearity of the linear predictor. In the **response scale**, the mean $\mathbf{\mu}$ is back transformed by the inverse link and respects the support of the distribution.

The link function is applied to the expected value ($E(\mathbf{y})$), and not to the observations. Transformation of the observations also effect the error, while link functions only affect the parameters controlling the expected value.

Example of link functions:

| Link Function | Equation | Use | Why |
|:----------------:|:----------------:|:----------------:|:----------------:|
| **Identity Link** | $g(\mu) = \mu$ | Normal dist. | $E(\mathbf{y})$ can take any real value ($-\infty, \; +\infty$) |
| **Logit Link** | $g(\mathbf{\mu}) = log(\frac{\mu}{1-\mu})$ | Logistic, Beta, Binomial dist. | $E(\mathbf{y})$ can take any values between 0 and 1. Maps $(0, \; 1) \rightarrow (-\infty, \; +\infty)$ |
| **Log Link** | $g(\mu) = log(\mu)$ | Poisson, Gamma dist. | $E(\mathbf{y})$ can take any positive values ($\mu > 0$) |

$$
\\[0.2in]
$$

#### Distributional assumption for the data

GLMMs support different distributions from the exponential family. Distributions from the exponential family share common structure, but are relatively different among themselves.

-   **Assumption**: Something you take as true about your data or about the process that generated it!

Important distributions to know are:

-   **For continuous data**: Normal, t, Gamma, Beta.

-   **For discrete data**: Binomial, Poisson, Negative Binomial.

$$
\\[0.2in]
$$

#### **Normal distribution**

$$
y \sim N(\mu, \; \sigma^2)
$$

$$
E(y) = \mu \\
Var(y) = \sigma^2
$$

**Support:**

$$
y \in (-\infty, \; +\infty)
$$

```{r, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(-10, 10, 0.1) 
plot(x, dnorm(x, 0, 1), type = "l", lwd = 2, ylab = "[x]", xlab = "x", ylim = c(0, 0.4)) 
lines(x, dnorm(x, 0, 4), lty = 2) 
lines(x, dnorm(x, 2, 1), lty = 3) 
legend("topright", legend = c("x ~ Normal(0,1)", "x ~ Normal(0,16)", "x ~ Normal(2, 1)"), lty = c(1,2,3), bty = "n")
```

$$
\\[0.05 in]
$$

#### **Student t distribution**

$$
y\sim t_v(\mu, \; \sigma^2)
$$

<center>

$E(y) = \mu$ for $v > 1$, otherwise undefined

$Var(y) = \frac{v}{v-2} \sigma^2$, otherwise undefined

</center>

**Support:**

$$
y \in (-\infty, \; +\infty)
$$

```{r, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(-10, 10, 0.1) 
plot(x, dt(x, 2), type = "l", lwd = 2, ylab = "[x]", xlab = "x", ylim = c(0, 0.4)) 
lines(x, dt(x, 100), lty = 2) 
lines(x, dt(x, 2, 2), lty = 3) 
legend("topright", legend = c("x ~ t(2)", "x ~ t(100)", "x ~ t(2, 2)"), lty = c(1,2,3), bty = "n")
```

$$
\\[0.05 in]
$$

#### **Gamma distribution**

$$
y \sim Gamma(\alpha, \; \beta)
$$

$$
E(y) = \frac{\alpha}{\beta} \\
Var(y) = \frac{\alpha}{\beta^2}
$$

**Support:**

$$
y \in (0, \; +\infty)
$$

```{r, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 12.5, 0.1) 
plot(x, dgamma(x, 2, 2), type = "l", lwd = 2, ylab = "[x]", xlab = "x", ylim = c(0, 1.5)) 
lines(x, dgamma(x, 10, 2), lty = 2) 
lines(x, dgamma(x, 3, 5), lty = 3) 
legend("topright", legend = c("x ~ Gamma(2, 2)", "x ~ Gamma(10, 2)", "x ~ Gamma(3, 5)"), lty = c(1,2,3), bty = "n")
```

$$
\\[0.05 in]
$$

#### **Beta distribution**

$$
y \sim Beta(\alpha, \; \beta)
$$

$$
E(y) = \frac{\alpha}{\alpha + \beta} \\
Var(y) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}
$$

**Support**:

$$
y \in (0, \; 1)
$$

```{r, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0.01, 0.99, 0.01) 
plot(x, dbeta(x, 0.5, 0.5), type = "l", lwd = 2, ylab = "[x]", xlab = "x", ylim = c(0, 3)) 
lines(x, dbeta(x, 1, 1), lty = 2) 
lines(x, dbeta(x, 4, 2), lty = 3) 
legend("top", legend = c("x ~ Beta(0.5, 0.5)", "x ~ Beta(1, 1)", "x ~ Beta(4, 2)"), lty = c(1,2,3), bty = "n")
```

$$
\\[0.05 in]
$$

#### **Poisson distribution**

$$
y \sim Poisson(\lambda)
$$

$$
E(y) = \lambda \\
Var(y) = \lambda
$$

**Support:**

$$
y \in (0, 1, 2, ..., +\infty)
$$

-   Model the number of events occurring in a fixed interval of time/space given a rate of occurrence ($\lambda$).

```{r, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 17, 1)
off.set <- 0.2
plot(x, dpois(x, 2), type = "h", lwd = 6, ylab = "[x]", xlab = "x", ylim = c(0, 0.3), col = adjustcolor("#C3C3C3", alpha.f = 1)) 
lines(x + off.set, dpois(x, 6), type = "h", lwd = 6, col = adjustcolor("#757575", alpha.f = 1)) 
lines(x + 2*off.set, dpois(x, 10), type = "h", lwd = 6, col = adjustcolor("#1E1E1E", alpha.f = 1))  
legend("topright", legend = c("x ~ Pois(2)", "x ~ Pois(6)", "x ~ Pois(10)"), lty = c(1, 1, 1), lwd = 6, col = c("#C3C3C3", "#757575", "#1E1E1E"), bty = "n")
```

$$
\\[0.05 in]
$$

#### Binomial distribution

$$
y \sim Binomial(n, p)
$$

$$
E(y) = np \\
Var(y) = np(1-p)
$$

**Support**:

$$
y \in (0, 1, ..., n) \\
n \in (1, 2, ..., +\infty) \\
p \in (0, 1)
$$

-   Model the number of successes in a fixed number of independent trials ($n$) with a given probability of success ($p$).

```{r, echo=FALSE, warning=FALSE, message=FALSE}
x <- seq(0, 11, 1)
off.set <- 0.2
plot(x, dbinom(x, 10, 0.2), type = "h", lwd = 6, ylab = "[x]", xlab = "x", ylim = c(0, 0.5), col = adjustcolor("#C3C3C3", alpha.f = 1)) 
lines(x + off.set, dbinom(x, 10, 0.5), type = "h", lwd = 6, col = adjustcolor("#757575", alpha.f = 1)) 
lines(x + 2*off.set,  dbinom(x, 10, 0.8), type = "h", lwd = 6, col = adjustcolor("#1E1E1E", alpha.f = 1))  
legend("topright", legend = c("x ~ Binom(10, 0.2)", "x ~ Binom(10, 0.5)", "x ~ Binom(10, 0.8)"), lty = c(1, 1, 1), lwd = 6, col = c("#C3C3C3", "#757575", "#1E1E1E"), bty = "n")
```

$$
\\[0.2 in]
$$

## Working with GLMMs

1.  Define a distribution that matches $y$.
2.  Define the linear predictor (fixed and random effects) $\eta$.
3.  Define the link function that connects $E(y)$ of the assume distribution and the linear predictor $\eta$.

$$
\\[0.2 in]
$$

## Packages

```{r, message=FALSE, warning=FALSE}
library(emmeans)
library(multcomp)
library(car)
library(DHARMa)
library(glmmTMB)
library(agridat)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readxl)
library(tidyverse)

# Severity data - Example 1
d1 <- read_excel("C:/Users/cdiasda/OneDrive - Kansas State University/Desktop/Doutorado/Laboratory and research/Workshops/Mixed Models Workshop/USA041.xlsx")
colnames(d1) <- c("fungicide", "block", "severity_o")
d1 <- transform(d1, fungicide = as.factor(fungicide), block = as.factor(block))
d1$severity <- d1$severity_o/100

# Germination data - Example 2
d2 <- crowder.seeds
d2$gen <- as.factor(d2$gen)
d2$extract <- as.factor(d2$extract)

# Count data - Example 3
d3 <- data.frame(
  species = c(rep("a", 10), rep("b", 10), rep("c", 10)),
  block = c(seq(1, 10, by = 1), seq(1, 10, by = 1), seq(1, 10, by = 1)),
  n_count = c(rpois(10, 3), rpois(10, 7), rpois(10, 12))
)
```

## Example I - Disease Severity

In this example we will evaluate disease severity. The data was generated from a designed experiment on fungicide efficacy to manage the disease know as yellow rust on wheat. The experimental design was a Randomized Complete Block Design (RCBD). Severity refers to how much an specific plant organ is affected by a given disease. In this case it refers to the leaf area covered by yellow rust lesions, also know as pustules.

**Data**

```{r, warning=FALSE, message=FALSE}
print(d1)
```

1.  Define a distribution that matches $y$.

$$ y \sim Beta(\mu, \; \phi) $$

-   Why?

    -   Severity is a percentage: 0 - 100%, in proportion: 0 - 1.

    -   The support from the Beta distribution perfectly matches our response.

    -   Remember the support for the Beta distribution:

$$
y \in (0, \; 1)
$$

2.  Define a linear predictor $\eta$.

$$ \eta_{ij} = \mu_0 + t_i + u_j $$

-   Where:

    -   $\mu_0$ represents the overall/gran mean.
    -   $t_i$ is the parameter for the effect of treatment, in this case, fungicides - **Fixed effect.**
    -   $u_j$ is the parameter for the effect of block - **Random effect.**

3.  Define the link function that connects $E(y)$ of the assume distribution and the linear predictor $\eta$.

<center>**Logit link**</center>

$$ g(\mu) = \eta = logit(\mu)$$

-   Why?

    -   Logit links $(-\infty, \; +\infty)$ to $(0, \; 1)$, that is our desired scale.

#### **Analysis of severity**

**Model**

$$
y_{ij}|u_j \sim Beta(\mu_{ij}, \; \phi) \\ logit(\mu_{ij}) = \eta_{ij} = \mu_0 + t_i + u_j \\ u_j \sim N(0, \sigma^2_u)
$$

**Fitting the model**

```{r, warning=FALSE, message=FALSE}
m1 <- glmmTMB(severity ~ fungicide + (1|block), family = beta_family(link = "logit"), data = d1)
summary(m1)
```

**Checking the model - Residuals**

```{r, warning=FALSE, message=FALSE}
res_sim1 <- simulateResiduals(m1, plot = TRUE)
```

**What are we checking here?**

Residuals: "The footprint of the fitted model" - How much model predictions deviate from observations.

$$ \epsilon_i = y_i - \hat{y}_i $$

For linear mixed models our residuals are assumed to be normally distributed and with constant variance ($\epsilon \sim i.i.d. N(0, \; \sigma^2_\epsilon)$). In models with different distributional assumptions, residuals do not follow a known distribution, which makes it tricky to check models in a similar way we did before.

**What is each test doing?**

-   QQ-Plot

    -   Compare the quantiles[^1] of two distributions, if they are similar, we expect a straight diagonal line. In this case, in the y-axis we have the quantiles of our residuals and in the x-axis the quantiles of a standard normal distribution.

    -   Kolmogorov-Smirnov test: Test for uniformity against a uniform distribution - $Uniform(0, \; 1)$.

    -   Dispersion test: Variance in the observations vs. Variance on the simulations.

    -   Outlier test: Residual values of 0 or 1. Test if the number of outliers is appropriate to the size of the data. Does not quantify the amount of outliers.

-   Residual vs Predicted

    -   Show whether we have homocedasticity of the variances (constant variance across predicted values) or not. Usually, if higher or lower predicted values have higher or lower variance, the plot will present a "funnel" shape. Ideally the plot is a random scatter of points.

    -   Fit smoothed splines in three points of the quantile residuals, 0.25, 0.50, and 0.75. Test whether these lines are flat or have trends. For a random scatter, we do not expect to see trends at any of these points, this would indicate heterocedasticity.

    -   It also indicated when outliers are detected by producing a red asterisk.

[^1]: Quantiles divide a dataset into equal-sized subsets, helping to understand more about the distribution of the data.

**ANOVA**

```{r, warning=FALSE, message=FALSE}
Anova(m1)
```

**Post-hoc test - Mean comparisons**

```{r, warning=FALSE, message=FALSE}
means1 <- emmeans(m1, ~fungicide, type = "response")
cld(means1, Letters = letters)
```

### What if we use a less appropriate assumption for severity?

Let's check Gamma!

```{r, warning=FALSE, message=FALSE}
m1_2 <- glmmTMB(severity_o ~ fungicide + (1|block), family = Gamma(link = "log"), data = d1)
res_sim1_2 <- simulateResiduals(m1_2, plot = TRUE)
means1_2 <- emmeans(m1_2, ~fungicide, type = "response")
cld(means1_2, Letters = letters)
```

## Example 2 - Seed germination

The following data arise from an experiment studying germination of Orobanche seeds [(Crowder, 1978)](https://www.jstor.org/stable/2346223?origin=crossref&seq=1).\
The data indicate the total number of seeds ($n$) and the number of germinated seeds ($germ$). The experiment was conducted as a completely randomized design with a 2 x 2 factorial treatment structure for type of extract (bean or cucumber) and extract concentration (o75 and o73).

**Data**

```{r, warning=FALSE, message=FALSE}
print(d2)
```

1.  Define a distribution that matches $y$.

$$ y \sim Binomial(n, \; p) $$

-   Why?

    -   We have the number of trials $n$

    -   For each trial we have the number of successes $germ$

    -   Remember the support for the Binomial distribution:

$$ y \in (1, 2, ..., n) $$

2.  Define a linear predictor $\eta$.

$$ \eta_{ij} = \mu_0 + ex_i + gen_j + (ex*gen)_{ij}$$

-   Where:

    -   $\mu_0$ represents the overall/gran mean.
    -   $ex_i$ is the parameter for the effect of extract - **Fixed effect.**
    -   $gen_j$ is the parameter for the effect of dilution - **Fixed effect.**
    -   $(ex*gen)_{ij}$ is the parameter for the effect of the interaction between the extract and the dilution.

3.  Define the link function that connects $E(y)$ of the assume distribution and the linear predictor $\eta$.

<center>**Logit link**</center>

$$ g(p) = \eta = logit(p)$$

-   Why?

    -   Logit links $(-\infty, \; +\infty)$ to $(0, \; 1)$, that is our desired scale.

**Model**

$$ y_{ij}|u_j \sim Binomial(n, \; p) \\ logit(p) = \eta_{ij} = \mu_0 + t_i + u_j \\ u_j \sim N(0, \sigma^2_u) $$

**Fitting the model**

```{r, warning=FALSE, message=FALSE}
m2 <- glmmTMB(cbind(germ, n-germ) ~ extract*gen, family = binomial(link = "logit"), data = d2)
summary(m2)
```

-   Different ways to fit the model with Binomial distribution - Depends on what you have.

    -   **Important point:** Remember what we are modeling - Probability of success: $p$.

    -   For **Binomial distribution:**

        -   **Case 1:** $y$ = Counts of success with know trials - Success and Failures = (success, trials - success).

        -   **Case 2:** $y$ = Counts of successes with weights - success/trials and weights = trials.

            -   Only if proportions come from counts!

        -   **Case 3:** A special case: $y$ = Binary outcomes (0/1 per observation) - Binomial with n = 1 - **Logistic regression**.

    -   For **Beta distribution**:

        -   **Case 4:** $y$ = Proportions not from counts (between 0 and 1) - No trial number, proportion only

            -   Do not use Binomial distribution here!

```{r, eval=FALSE, warning=FALSE, message=FALSE}
# Case 1
glmmTMB(cbind(germ, n-germ) ~ extract*gen, family = binomial(link = "logit"), data = data)

# Case 2
glmmTMB(germ/n ~ extract*gen, family = binomial(link = "logit"), weights = n, data = data)

# Case 3 - If germinated = yes (1) / no (0)
glmmTMB(germinated ~ extract*gen, family = binomial(link = "logit"), data = data)

# Case 4 - From example 1 - Severity: Proportion of area damaged by disease
glmmTMB(severity ~ fungicide + (1|block), family = beta_family(link = "logit"), data = data)
```

For the logistic regression:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
x <- c(0, 1)
plot(x, dbinom(x, 1, 0.6), main = "x ~ Binom(1, 0.6)", type = "h", lwd = 6, ylab = "[x]", xlab = "x", ylim = c(0, 1), col = adjustcolor("#1E1E1E"))
```

**Checking the model**

```{r, warning=FALSE, message=FALSE}
res_sim2 <- simulateResiduals(m2, plot = TRUE)
```

**ANOVA**

```{r, message=FALSE, warning=FALSE}
Anova(m2)
```

**Marginal means**

```{r, warning=FALSE, message=FALSE}
emmeans(m2, ~extract*gen, type = "response")
```

## Example 3 - Population count - SIMULATED DATA

The following data is simulated to mirror results from a randomized complete block experiment in which the number of individuals from a given pest insect are counted in three different plant species.

**Data**

```{r, message=FALSE, warning=FALSE}
print(d3)
```

1.  Define a distribution that matches $y$.

$$ y \sim Poisson(\lambda) $$

-   Why?

    -   We have counts of events - Number of individuals per pot.

    -   Remember the support for the Poisson distribution:

$$ y \in (1, 2, ..., +\infty) $$

2.  Define a linear predictor $\eta$.

$$ \eta_{ij} = \mu_0 + sp_i + u_j$$

-   Where:

    -   $\mu_0$ represents the overall/gran mean.
    -   $sp_i$ represents the effect of the ith species - **Fixed effect**.
    -   $u_j$ represents the effect of the jth block - **Random effect**.

3.  Define the link function that connects $E(y)$ of the assume distribution and the linear predictor $\eta$.

<center>**Log link**</center>

$$ g(\lambda) = \eta = log(\lambda)$$

-   Why?

    -   Log ensures that the expected value will be positive.
    -   Remember:

$$
\lambda \in \mathbb{R}^+
$$

**Model**

$$ y_{ij}|u_j \sim Poisson(\lambda_{ij}) \\ logit(\lambda_{ij}) = \eta_{ij} = \mu_0 + t_i + u_j \\ u_j \sim N(0, \sigma^2_u) $$

**Fitting the model**

```{r, message=FALSE, warning=FALSE}
m3 <- glmmTMB(n_count ~ species + (1|block), family = poisson(link = "log"), data = d3)
summary(m3)
```

-   Always check for over dispersion! In Poisson, variance increases with the mean, once they are both controlled by $\lambda$. If this is not true, Negative Binomial distribution might be an alternative option.

**Checking the model**

```{r, message=FALSE, warning=FALSE}
sim_res3 <- simulateResiduals(m3, plot = TRUE)
```

**ANOVA**

```{r, message=FALSE, warning=FALSE}
Anova(m3)
```

**Marginal means**

```{r, message=FALSE, warning=FALSE}
emmeans(m3, ~species, type = "response")
```

## Hierarchical models

**Why are mixed models sometimes called 'hierarchical' or 'multilevel' models?**

What is an hierarchical model?

$$ y_{ij}|u_j \sim Beta(\mu_{ij}, \; \phi) \\ logit(\mu_{ij}) = \eta_{ij} = \mu_0 + t_i + u_j \\ u_j \sim N(0, \sigma^2_u) $$

If we use the Beta regression model for disease severity as an example, we can see:

**Data model:** The conditional distribution we are assuming for $y$.

$$
y_{ij}|u_j \sim Beta(\mu_{ij}, \; \phi)
$$

**Process model:** The functions that shapes the expected value/mean $\mu$, and the link function connecting it to the boundaries of the assumed distribution.

$$
logit(\mu_{ij}) = \eta_{ij} = \mu_0 + t_i + u_j
$$

**Parameter model:** The random effects structure, the distribution of the random effects that capture the variation among groups.

$$
u_j \sim N(0, \; \sigma_u^2)
$$

**Linear Mixed Effects are also hierarchical models**, but with two levels:

-   The data model - $y_{ij}|u_j \sim N(\mu_{ij}, \; \sigma^2)$.

-   And the parameter model - $u_j \sim N(0, \; \sigma^2)$.

**But what it actually means?**

Let's look at a split-plot design example:

**The model:**

$$
y_{ijkl}|u_k, \; w_l \sim Beta(\mu_{ijkl}, \; \phi) \\
logit(\mu_{ijkl}) = \eta_{ijkl} = \mu_0 + f_i + v_j + u_k + w_l \\
u_k \sim N(0, \; \sigma_u^2) \\
w_l \sim N(0, \; \sigma_w^2)
$$

Where:

-   $\mu_0$ is the overall mean.

-   $f_i$ is the fixed effect of fungicide applied to the whole plot.

-   $v_j$ is the fixed effect of variety applied to the subplot level.

-   $u_k$ is the random effect of the block.

-   $w_l$ is the random effect of the whole plot level, that comes from $u_k*f_i$.

-   The subplot level is nested with residuals, which is parameterized by $\phi$.
